---
title: Converting single-cell data structures to/from Bioconductor and Python
author: 
- name: Luke Zappia
  email: luke@lazappi.id.au
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "Revised: 16 June 2020"
output:
  BiocStyle::html_document:
    toc_float: true
package: zellkonverter 
vignette: >
  %\VignetteIndexEntry{Converting to/from AnnData to SingleCellExperiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
library(BiocStyle)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Overview

This package provides a lightweight interface between the Bioconductor `SingleCellExperiment` data structure
and the Python `AnnData`-based single-cell analysis environment.
The idea is to enable users and developers to easily move data between these frameworks 
to construct a multi-language analysis pipeline across R/Bioconductor and Python.

# Writing H5AD files 

Users can use the `readH5AD()` function to read a `SingleCellExperiment` from a H5AD file.
This can be manipulated in the usual way as described in the `r Biocpkg("SingleCellExperiment")` documentation.

```{r}
library(zellkonverter)

# Obtaining an example file.
example.file <- system.file("extdata", "krumsiek11.h5ad", package="zellkonverter")
readH5AD(example.file)
```

We can also write a `SingleCellExperiment` to a H5AD file with the `writeH5AD()` function.
This is demonstrated below on the classic Zeisel mouse brain dataset from the `r Biocpkg("scRNAseq")` package;
the subsequent file can then be directly used in compatible Python-based analysis frameworks.

```{r}
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()
loc <- tempfile(pattern=".h5ad")
writeH5AD(sce.zeisel, file=loc)
```

# Creating `AnnData` objects 

Developers and power users who control their Python environments can 
directly convert between `SingleCellExperiment` and [`AnnData` objects](https://anndata.readthedocs.io/en/stable/)
using the `SCE2AnnData()` and `AnnData2SCE()` utilities.
These functions expect that `r Biocpkg("reticulate")` has already been loaded
along with an appropriate version of the [_AnnData_](https://pypi.org/project/anndata/) package.
We suggest using the `r Biocpkg("basilisk")` package to set up the Python environment before using these functions.

```{r}
library(basilisk)
library(scRNAseq)
seger <- SegerstolpePancreasData()
roundtrip <- basiliskRun(env=zellkonverter:::anndata_env, fun=function(sce) {
     # Convert SCE to AnnData:
     ad <- SCE2AnnData(sce)

     # Maybe do some work in Python on 'ad':
     # BLAH BLAH BLAH

     # Convert back to an SCE:
     AnnData2SCE(ad)
}, sce=seger)
```

Package developers can guarantee that they are using the same versions of Python packages as `r Biocpkg("zellconverter")`
by using the `.AnnDataDependencies` variable to set up their Python environments.

```{r}
.AnnDataDependencies
```

# Session information

```{r}
sessionInfo()
```
